---
title: "Untitled"
output: html_document
date: "2023-08-17"
---

```{r setup, include=FALSE}
library(tidyverse)

library(ggnetwork)
# only for generating example graph
library(network)

knitr::opts_chunk$set(echo = TRUE)
```

## Plotting Graphs with `ggnetwork`

[ggnetwork vignette](https://cran.r-project.org/web/packages/ggnetwork/vignettes/ggnetwork.html)

---

Install packages if necessary

```{r}
# intall the package if not already installed
# install.packages('ggnetwork')

# it might also be helpful to install packages for work with graph data
# install.packages(c('network', 'sna', 'igraph'))
```
<br>

### Example Graph Data

```{r}
# Let’s define a small random graph to illustrate each component of ggnetwork:
n <- network::network(sna::rgraph(10, tprob = 0.2), directed = FALSE)

# Let’s now add categorical and continuous attributes for both edges and 
# vertices. We’ll start with nodes, adding a categorical vertex attribute 
# called "family", which is set to either "a", "b" or "c", and a continuous 
# vertex attribute called "importance", which is set to either 1, 2 or 3.
n %v% "family" <- sample(letters[1:3], 10, replace = TRUE)
n %v% "importance" <- sample(1:3, 10, replace = TRUE)

# We now add a categorical edge attribute called "type", which is set to either 
# "x", "y" or "z", and a continuous vertex attribute called "day", which is set 
# to either 1, 2 or 3.
e <- network.edgecount(n)
set.edge.attribute(n, "type", sample(letters[24:26], e, replace = TRUE))
set.edge.attribute(n, "day", sample(1:3, e, replace = TRUE))
```

<br>

### Plotting

#### Flattening Graph to Table

```{r}
ggnetwork(n)
```

#### Plotting

Basic example. `ggnetwork` comes with fortify methods for graphs of class `network` and `igraph` so you can plot those objects directly with `ggplot()`, but may have to pipe through `ggnetwork()` first if the fortify isn't working properly or you don't explicitly load the package

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey50") +
  theme_blank()

# may have to pipe to ggnetwork() first if ggplot() isn't working with the 
# graph time correctly
n %>%
  ggnetwork() %>%
ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(aes(linetype = type), color = "grey50") +
  theme_blank() 
```

With more options, like points to represent the nodes as well as text about the graph with coloring based on features in the graph

```{r}
ggplot(n, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_edges(color = "black") +
  geom_nodelabel_repel(aes(color = family, label = LETTERS[ vertex.names ]),
                       fontface = "bold", box.padding = unit(1, "lines")) +
  geom_nodes(color = "black", size = 8) +
  theme_blank()
```





<br>
