---
title: ""
output: html_document
date: "2023-02-07"
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(viridis)
library(plotly)
library(ggbeeswarm)

knitr::opts_chunk$set(echo = TRUE)
```

## Setting Mouseover/Tooltip Text in `plotly` Interactive Plots

By default, `plotly::ggplotly()` tooltip/mouseover text shows the values of the variables for that point, or for distributions shows the summary values of the distribution. You can specify the text yourself either for formatting purposes or to include data that isn't otherwise displayed on the plot.

### Scatterplot

```{r}
### defaults
diamonds %>%
  # filtered for color only because there are a lot of data points and the 
  # interactive plot was sluggish
  filter(color == 'J') %>%
ggplot(aes(x = carat, y = depth, color = clarity)) +
  scale_color_viridis(discrete = T) +
  geom_point() +
  theme_bw() -> scatter_default
ggplotly(scatter_default)
```

```{r}
### setting the text manually
# adding to the end
diamonds %>%
  filter(color == 'J') %>%
  # you can just add on another variable to the text geom and it will be added
  # 
  mutate(text = paste0('cut = ', cut)) %>%
ggplot(aes(x = carat, y = depth, color = clarity, text = text)) +
  scale_color_viridis(discrete = T) +
  geom_point() +
  theme_bw() -> scatter_manual1
# default tooltip is set to all variables
ggplotly(scatter_manual1, tooltip = 'all')

# specifying the whole thing yourself
diamonds %>%
  filter(color == 'J') %>%
  # don't forget the \n at the end or everything will be on one line
  mutate(text = paste0('carat = ', carat, '\n',
                       'depth = ', depth, '\n',
                       'clarity = ', clarity, '\n',
                       'cut = ', cut)) %>%
ggplot(aes(x = carat, y = depth, color = clarity, text = text)) +
  scale_color_viridis(discrete = T) +
  geom_point() +
  theme_bw() -> scatter_manual2
# have to set tooltip = 'text' so you will get only your manual specification
ggplotly(scatter_manual2, tooltip = 'text')
```

### Boxplot

NOTE: This doesn't work

```{r}
### default
diamonds %>%
  filter(color == 'J') %>%
ggplot(aes(x = clarity, y = depth, fill = cut)) +
  geom_boxplot() +
  scale_fill_viridis(discrete = T) +
  theme_bw() -> boxplot_default
ggplotly(boxplot_default)

### with specified text
diamonds %>%
  filter(color == 'J') %>%
  mutate(text = cut) %>%
ggplot(aes(x = clarity, y = depth, fill = cut)) +
  geom_boxplot() +
  scale_fill_viridis(discrete = T) +
  theme_bw() -> boxplot_labeled
ggplotly(boxplot_labeled, hoverinfo = 'text') %>% 
  # style(text = 'text')
  layout(hoverinfo = 'none')
```



<br><br>

