---
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(conflicted)
library(viridis)
library(magrittr)
library(UpSetR)

filter <- dplyr::filter

knitr::opts_chunk$set(echo = TRUE)
```

## UpSetR Tutorial

UpSet plots are a better way to visualize more than two intersecting sets

-

UpSetR documentation here: <https://github.com/hms-dbmi/UpSetR>

### Set Up

Fake methylation data! for the the tutorial

```{r}
tibble(sma_id = 'SmaI_mm9',
       site_number = rep(sample(1000), 3),
       sample = factor(rep(c('cond1', 'cond2', 'cond3'), each = 1000),
                       levels = c('cond1', 'cond2', 'cond3')),
       meth_diff = c(runif(n = 3000, min = -100, max = 100)),
       pvalue = rnorm(3000)) %>%
  unite(site_id, sma_id, site_number, sep = '_') -> fake_meth_data
```

### Wrangle Data

The UpSetR function `upset()` needs data to be in a matrix format where every set is its own column and the column contains 1 and 0 where 1 = the thing is in the set and 0 = the thing is not in the set

```{r}
fake_meth_data %>%
# filter for significant sites by pvalue
  filter(pvalue < 0.05) %>%
# take the pvalue column away before spreading
  select(-pvalue) %>%
# spread the data into a wide format for the upset function
  spread(sample, meth_diff) %>%
# replace the NAs with 0 because those sites weren't observed
  replace(is.na(.), 0) %>%
# annotate columns with 0/1 based on whether they "changed" more than 5% and the
# direction of the change
  mutate(cond1_hypo = ifelse(cond1 <= -5, 1, 0),
         cond1_hyper = ifelse(cond1 >= 5, 1, 0),
         cond2_hypo = ifelse(cond2 <= -5, 1, 0),
         cond2_hyper = ifelse(cond2 >= 5, 1, 0),
         cond3_hypo = ifelse(cond3 <= -5, 1, 0),
         cond3_hyper = ifelse(cond3 >= 5, 1, 0)) %>%
# drop the columns with the methylation data (not 100% necessary)
  select(site_id, cond1_hypo:cond3_hyper) %>%
# the table MUST be a data.frame for the upset() function
  as.data.frame() -> upset_tbl
```

### Plot UpSet Plots!

The absolute basics: the table and the columns. All you need to give the function is your table of 1s and 0s and tell it which columns to plot.

```{r}
# hypomethylated sites
upset(upset_tbl,
# list the columns you want to plot
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'))

# hypermethylated sites
upset(upset_tbl,
# list the columns you want to plot
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'))
```

#### Reorder the bars

The optional argument `order.by` lets you set the order of the bars. When you specify "degree", the bars are ordered by the number of intersections, starting with the highest number on the left, in this case 3, and going down to unique things on the right.

```{r}
# hypomethylated sites
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      order.by = 'degree')

# hypermethylated sites
upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      order.by = 'degree')
```

When you specify "freq", the bars are ordered starting with the largest set on the left and then in descending order by set size.

```{r}
# hypomethylated sites
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      order.by = 'freq')

# hypermethylated sites
upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      order.by = 'freq')
```

You can also combine both arguments.

```{r}
# degree first, freq second
# hypomethylated sites
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      order.by = c('degree', 'freq'))

# hypermethylated sites
upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      order.by = c('degree', 'freq'))

###

### freq first, degree second
# hypomethylated sites
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      order.by = c('freq', 'degree'))

# hypermethylated sites
upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      order.by = c('freq', 'degree'))
```

#### Change the Appearance of the UpSet Plots

Change the main bar colors. You can make them a single color, or give a vector of colors for each individual bar

```{r}
### one main bar color
# hypomethylated sites
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      main.bar.color = 'deepskyblue3')

# hypermethylated sites
upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      main.bar.color = 'firebrick3')

### each bar is a different color
upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      main.bar.color = c('firebrick3', 'orange', 'lightgoldenrod1', 
                         'green4', 'blue', 'midnightblue', 'violet'))
```

Set the color of the intersection points

```{r}
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      matrix.color = c('orange'))
```

Set the color of the "Set Size" bar plot. Can give it one color, or can give it a vector of colors so each set is different.

```{r}
# one color
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      sets.bar.color = c('green4'))

# vector of colors for all bars
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      sets.bar.color = c('magenta4', 'sienna2', 'green4'))
```

Set labels

```{r}
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      mainbar.y.label = 'This sets the y label!',
      sets.x.label = 'This sets the x label!')
```

#### All together now!

```{r}
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'deepskyblue3',
      mainbar.y.label = '',
      sets.bar.color = c('magenta4', 'sienna2', 'green4'),
      text.scale = c(1, 2, 1, 1, 2, 2))

upset(upset_tbl,
      sets = c('cond1_hyper', 'cond2_hyper', 'cond3_hyper'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'firebrick3',
      mainbar.y.label = '',
      sets.bar.color = c('magenta4', 'sienna2', 'green4'),
      text.scale = c(1, 2, 1, 1, 2, 2))
```

### Saving the UpSet Plot

Because an UpSet plot isn't a normal ggplot object, if you want to use `ggsave()`, you have to wrap the UpSet plot function in `ggsave()`. Example in the chunk below. To try saving, uncomment the `ggsave` lines and run the chunk.

```{r}
#ggsave('practice_saving_upset.png',
upset(upset_tbl,
      sets = c('cond1_hypo', 'cond2_hypo', 'cond3_hypo'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'deepskyblue3',
      mainbar.y.label = '',
      sets.bar.color = c('magenta4', 'sienna2', 'green4'),
      text.scale = c(1, 2, 1, 1, 2, 2))
#)
```






