
---
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(conflicted)

filter <- dplyr::filter

knitr::opts_chunk$set(echo = TRUE)
```

## Volcano Plot

### Fake Data

You need the `TeachingDemos` package to simulate pvalues, so if you don't have it installed, uncomment the code in the chunk below and install it.

```{r}
#install.packages('TeachingDemos')
```

Simulate pvalues and methylation differences

```{r}
tibble(pvalue = TeachingDemos::Pvalue.norm.sim(n = 10000),
       meth_diff = rnorm(10000, mean = 0, sd = 15)) -> fake_data
```

### Wrangle Data

Wrangle the fake methylation data for plotting

```{r}
fake_data %>%
  mutate(logpvalue = -log10(pvalue),
         sig = case_when(pvalue < 0.05 & meth_diff < -5 ~ 'hypo',
                         pvalue < 0.05 & meth_diff > 5 ~ 'hyper',
                         TRUE ~ 'not_sig')) -> wrangled_fake_data
```

### Plot

Get labels for plot. Labels with have the count of "significant" sites and the total percentage those "significant" sites are

```{r}
wrangled_fake_data %>% 
# count the number of significant hyper- and hypo- methylated sites
  group_by(sig) %>% 
  count() %>% 
  ungroup() %>%
  filter(sig != 'not_sig') %>%
# column sets location on the x-axis; needs to have the same name as the x-axis 
# column name in the main tablece
  mutate(meth_diff = c(30, -30), 
# column sets location on the y-axis; needs to have the same name as the y-axis 
# column name in the main table
         logpvalue = 3.5,
# calculate what percentage the counts are and round to the nearest 1 decimal 
# place and combine it with the count to make the labels
         label = paste0(n, 
                        ', ', 
                        round(((n / nrow(wrangled_fake_data)) * 100), 1), 
                        '%')) -> labels
```

Plot the volcano plot. (**NOTE:** Because the data is re-simulated every time the document is run, some parts of the plot might not look the best, like the placement of labels or the symmetry of the x-axis. This can be adjusted for your own non-simulated data.)

```{r, fig.width = 6, fig.height = 6}
ggplot(wrangled_fake_data, aes(x = meth_diff, y = logpvalue)) +
  geom_point(aes(color = sig), size = 2) +
  scale_color_manual(values = c('darkgoldenrod3', 'deepskyblue4', 'gray60')) +
  geom_text(data = labels, aes(label = label), size = 8) +
  geom_vline(xintercept = c(-5, 5), 
             color = 'gray40', 
             linetype = c('dashed')) +
  geom_hline(yintercept = -log10(0.05), 
             color = 'gray40', 
             linetype = 'dashed') +
  labs(x = 'Difference in % Methylation',
       y = '-Log10 PValue') +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')
```

