---
title: "`fgsea`"
output: html_document
date: "2023-01-17"
---

```{r setup, include=FALSE}
library(tidyverse)
# was having issues with my local machine finding gfortran; specifying the path
# was suggested to fix the issue, but there's probably some issue with my 
# gcc/gfortran install because this didn't fix the issue
# dyn.load('/opt/R/arm64/gfortran/lib/libgfortran.5.dylib')
library(fgsea)

### required but not explicitly loaded to improve speed and reduce namespace 
### conflicts
# library(reactome.db)
# library(viridis)
# library(KEGGREST)
# library(org.Mm.eg.db)
# library(msigdb)

knitr::opts_chunk$set(echo = TRUE)
```

## `fgsea`

Using the Fast Gene Set Enrichment `fgsea` package for pathway analysis in RNA-seq <https://bioconductor.org/packages/release/bioc/html/fgsea.html>

Databases demoed:
- Reactome
- KEGG
- MSigDB (Hallmark, C2CP pathways, and C5 pathways [GO])

---

#### Load the Example Data from the Package

This is a named list of mouse Entrez IDs (name) with their test statistics (value)

```{r}
data("exampleRanks")

exampleRanks %>% head()
```

<br>
---
---
<br>

## Using `fgsea` With Various Pathway Databases

### `fgsea` Using the Built-In `Reactome` Database

#### Test

Load the Reactome database using the built-in pathway functions

```{r}
### load built-in Reactome pathways
# when doing this you need to provide all the gene names in your dataset
reactome_pathways <- reactomePathways(names(exampleRanks))
```

Run the pathway analysis. You need to specify the list with the pathways, your vector with the genes and their significance, and, optionally, the number of permutations for the empirical p-value calculation. Should only need to specify the number of permutations if you have a small sample size and need to reduce from the default.

```{r}
### calculate the pathway enrichment using the example data and the reactome
### pathways loaded above
fgsea_res <- fgsea(pathways = reactome_pathways, 
                   stats = exampleRanks)
```

#### Results

Look at the results! The `fgsea()` function returns a `data.frame`, which makes it easy to use for plotting or other actions afterwards. The table contains information on how enriched each pathway is in the data and how significantly. You can check this information for yourself in the `fgsea()` function documentation anytime, but the columns are:

- **pathway**: name of the pathway as given in the pathway list
- **pval**: "significance" of the enrichment score calculated by a permutation test
- **padj**: p-value adjusted for multiple testing using Benjamini-Hotchberg (BH)
- **ES**: pathway enrichment score; positive for upregulated and negative for downregulated
- **NES**: enrichment score normalized by permutation test
- **nMoreExtreme**: the number of times a random gene set had a more extreme enrichment enrichment value in the permutations (out of 10,000 here because that's the number of permutations we specified)
- **size**: number of genes present in the data in the pathway
- **leadingEdge**: vector of genes that drive the enrichment

To get the genes out of the leadingEdge column, use the `tidyr::unnest()` function. Note that everything in the row that was already there gets duplicated for each gene name that was in the nested vector.

```{r}
# fgsea() returns a data.frame! YAY!!
fgsea_res %>% class()
# look at the data
fgsea_res %>% filter(pathway == "Cell Cycle")
# get the genes that are driving the enrichment score
fgsea_res %>%
  unnest(c(leadingEdge))
```

#### Visualize Results

If there's a particular pathway of interest, you can plot the cumulative enrichment curve for it using the pathways and data used to calculate the gene set enrichment, using `fgsea`'s `plotEnrichment()` function. It returns a `ggplot` object, so you can modify it using `ggplot` arguments somewhat.

```{r}
plotEnrichment(reactome_pathways[["Cell Cycle"]],
               exampleRanks) +
  labs(title = "Cell Cycle")
# The black lines on the x-axis, the rug, represent genes that are in the 
# pathway. It's easy to see in the pathway below where only 2 genes from the 
# data are in it.
plotEnrichment(reactome_pathways[["5-Phosphoribose 1-diphosphate biosynthesis"]],
               exampleRanks) +
  labs(title = "Cell Cycle")

# does return a ggplot plot, so can modify using standard ggplot arguments
plotEnrichment(reactome_pathways[["Cell Cycle"]],
               exampleRanks) %>% class()
```

Look at the top 10 most up- and down-regulated statistically significant pathways.

```{r}
fgsea_res %>% 
  as_tibble() %>% 
# filter for significant pathways
  filter(padj < 0.05) %>% 
# arrange by the normalized enrichment score
  arrange(NES) %>% 
# get the first and last 10 rows which will be the 10 most up- and down- 
# regulated pathways
  mutate(rownum = row_number()) %>%
  filter(rownum %in% c(1:10, (max(rownum - 9)):max(rownum))) %>%
# the pathway names can be long, so if they're over 20 characters, subset them
# otherwise use the whole name
  mutate(pathway_short = ifelse(nchar(pathway) <= 30, 
                                pathway, 
                                paste0(str_sub(pathway, start = 1, end = 27), 
                                       '...'))) %>%
ggplot(aes(x = reorder(pathway_short, NES), y = NES)) +
  geom_col(aes(fill = padj)) +
  viridis::scale_fill_viridis() +
  coord_flip() +
  labs(x = 'Pathway', 
       y = 'Normalized Enrichment Score (NES)', 
       fill = 'Q-Value') +
  theme_minimal(base_size = 16)
```

<br>

### KEGG

Get KEGG database info using the `KEGGREST` package <https://bioconductor.org/packages/release/bioc/html/KEGGREST.html>

```{r}
org <- KEGGREST::keggList("organism")
# find the abbreviation for mouse
org %>% as.data.frame() %>% filter(str_detect(tolower(species), 'mus'))
# NOTE: mouse is "mmu" and human is "hsa", so to change this code over to human
# you just need to swap mmu for hsa and switch to the human org.db

# simplest version to just load a named vector of (human/hsa) pathways
# KEGGREST::keggLink(target = "pathway", source = "hsa") -> kegg_pathways

### load all the mouse pathways needed to run pathway analysis
# The names in the vector of returned pathways are gene names in the format 
# "KEGG_SPECIES_ABBREVIATION:ENTREZ_ID. Transform the vector into a tibble, 
# remove the extra "mmu:" and get the ensembl and gene names (using 
# AnnotationDbi::mapIds()) which you'll probably want later for reference
KEGGREST::keggLink("pathway", "mmu") %>% 
  tibble(pathway = ., eg = sub("mmu:", "", names(.))) %>%
  mutate(symbol = AnnotationDbi::mapIds(org.Mm.eg.db::org.Mm.eg.db, eg, 
                                        "SYMBOL", "ENTREZID"),
        ensembl = AnnotationDbi::mapIds(org.Mm.eg.db::org.Mm.eg.db, eg, 
                                        "ENSEMBL", "ENTREZID")) -> kegg_paths


### load all the pathway descriptions which you'll want later to annotate
### the results using KEGGREST::keggList()
KEGGREST::keggList("pathway", "mmu") %>% 
    tibble(pathway = names(.), description = .) -> kegg_paths_description


### combine pathway identifiers with the de
left_join(kegg_paths, 
          kegg_paths_description, by = 'pathway') -> kegg_paths_anno


### format the pathway descriptions with their Entrez IDs for running fgsea,
### only keep pathways that contain a gene that was detected
hsa_kegg_anno %>%
  filter(eg %in% names(exampleRanks)) %>%
  select(pathway, eg) %>%
  unstack(eg ~ pathway) -> kegg_pathway_list
```

Run `fgsea` with the kegg pathways

```{r}
fgsea_res_kegg <- fgsea(pathways = kegg_pathway_list, 
                        stats = exampleRanks)

### look at the results
fgsea_res_kegg

### we got the pathway descriptions so you could join back and get an 
### human-intelligible name for the pathway
fgsea_res_kegg %>%
  left_join(distinct(select(kegg_paths_anno, pathway, description)), 
            by = 'pathway') %>%
  select(pathway, description, everything()) %>%
  # may also want to remove the species name at the end of every pathway 
  # description since it's redundant
  mutate(description = str_remove(description, " - Mus musculus \\(house mouse\\)"))
```

<br>

### MSigDB

#### Using the `msigdb` Package

<https://bioconductor.org/packages/release/data/experiment/html/msigdb.html>

```{r}
msigdb.mm <- msigdb::getMsigdb(org = 'mm', id = 'EZID')
```

```{r}
### subsetting the collections you want
msigdb::listCollections(msigdb.mm)

msigdb::listSubCollections(msigdb.mm)

# hallmark collections
msigdb::subsetCollection(msigdb.mm, 'h') %>%
# transforms into list format needed for fgsea
  GSEABase::geneIds() -> msigdb.mm_hallmarks

# gene ontology biological processes
msigdb::subsetCollection(msigdb.mm, 'c5', 'GO:BP') %>%
  GSEABase::geneIds()
```

Run `fgsea` with the kegg pathways

```{r}
fgsea_res_msigdb <- fgsea(pathways = msigdb.mm_hallmarks, 
                          stats = exampleRanks)

### look at the results
fgsea_res_msigdb
```

#### Using Files Directly from MSigDB

```{r cars}
fgsea::gmtPathways('~/Downloads/h.all.v2022.1.Hs.entrez.gmt') %>% head()
```





<br><br>

